Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-12-17 pkg/glmertree/paper/treatment_interactions.Rnw
Minor paper updates and updated simulation files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1054 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates. Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

@@

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}), GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we first introduce an artificial motivating dataset, which will be used to illustrate and explain treatment-subgroup interaction detection with GLM-based MOB and GLMM tree. In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. The comparison with GLM-based MOB allows for assessing the extent to which including random-effects estimation improves the performance of MOB trees. The comparison with GLMMs with pre-specified interactions will allow for assessing the extent to which using a tree-based method will improve detection of treatment-subgroup interactions. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research.

@@

We created a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the pos-ttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

The simulated dataset has 3 subgroups with differential treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment~1 is more effective than Treatment~2: the mean of the response variable is 7 for Treatment~1, and 11 for Treatment~2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment~1, and 9 for Treatment~2. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment~2 is more effective than Treatment~1: the mean value of the response variable is 12 for Treatment~1, and 7 for Treatment~2.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$. For simplicity we only employ a cluster-specific intercept here, but further random effects can be easily included in $z_i$. Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree. In Study I, we assessed the accuracy of both tree-based algorithms in datasets with treatment-subgroup interactions. In Study II, we assessed the Type-I error of both tree-based algorithms in datasets without treatment-subgroup interactions. In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

In all simulation studies, we varied the following data-generating parameters:

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
\item Two levels for the intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates.  
\end{enumerate}

In addition, each simulation study required a different design for main and interaction effects, which will be described in their respective sections below.

Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. Means for the other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

@@

In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$). Therefore, an extra level was added to the sixth facet of the \textit{General simulation design} described above. Also, to assess the effect of the size of the treatment-effect differences in nodes 3 and 7, an additional facet was added:

\begin{enumerate}
\item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.  
\item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

Again, 50 datasets were generated for each cell of the design. As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated. In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.

@@

To assess the accuracy of trees recovered by both algorithms, we inspected variables and values that were selected for partitioning. For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).

@@

The total number of nodes in estimated GLM and GLMM tree were calculated. A tree of size $>1$ was classified as a Type-I error. To identify the most important predictors of Type-I error, we used a GLMs with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors were further assessed with a graphical display.

@@

The graphical display in Figure~\ref{fig:xyplot_treesize_nointeract} indicates that GLMM tree has a Type-I error rate somewhat below the pre-specified $\alpha$ level under all circumstances. The same goes for GLM tree, whenever the random intercept has 0 variance, or the random intercept is not correlated to one of the partitioning covariates. However, when the random intercept is correlated to one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree. With large sample sizes and large variance of the random intercept, GLM tree will almost certainly create spurious splits.

@@

For datasets with purely continuous interactions, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$. Here, $\beta$ has a global value and no subscript, and comprised purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.

@@

\subsubsection{GLMMs with pre-specified interactions}
GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order interactions between all pairs of covariates $U_k$, and second-order interactions between all pairs of covariates $U_k$ and treatment. Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the GLMM. Predictions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM. Predictions for test observations were calculated using the refitted GLMMs.

\subsubsection{Assessment of performance} 
Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets. To identify the most important predictors of predictive accuracy, we used an ANOVA with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors were further assessed with a graphical display.

@@

GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .54 (SD = .43) for GLMMs. The most important predictors of predictive accuracy were algorithm type, sample size, the number of covariates, and type of interaction (continuous, piecewise or both). These effects were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.

As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions. Performance of both methods improves with increasing sample size. Furthermore, performance of GLMM tree is not affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous. This indicates that GLMM tree may be especially useful for exploratory purposes, where there are many potential moderator variables. Also, GLMM trees may provide simpler models, as the average number of fixed-effects predictor variables (main effects and interactions) evaluated for prediction with GLMMs was 12.30, whereas the average number of inner nodes that need to be evaluated for prediction with GLMM tree was only 3.38.

@@

In the current paper, we presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies. We performed simulation studies to assess the accuracy of GLMM tree in detecting treatment-subgroup interactions and illustrated the application of GLMM tree on an existing dataset.

@@

GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree. GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (i.e., Cohen's $d=.5$), and/or sample size was small (i.e., $ N < 1,000$). Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.

@@

Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided somewhat better accuracy, on average. When interactions were purely continuous, GLMMs outperformed GLMM trees, and when interactions were partly or purely piecewise, GLMM trees outperformed GLMMs. In addition, GLMM tree may have a clear advantage when there are a large number of potential moderator variables (i.e., $> 5$). Therefore, GLMM trees may be better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates. Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models. The trees in our simulations were limited to a maximum number of 7 inner nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.

@@

In the Application, we found GLMM tree to provide results that were more easily interpretable, and also more accurate than a GLM tree without random effects. In addition, to judge clinical relevance of the findings, we calculated node-specific effect sizes, as would often be done in RCTs or meta-analysis. Although we have limited ourselves to calculating Cohen's $d$ in the current paper, equivalent values of the success rate difference or the number needed to treat can be calculated, but this would involve additional distributional assumptions \cite{KraeyKupf06}. Node-specific effect sizes can also be used to prune trees, when a researcher prefers to have a final tree which is based on statistical as well as clinical significance. A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process.

>>>>

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates. Compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy, on average. We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

@@

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}), GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we first introduce GLM-based MOB and the GLMM tree algorithm. In the Simulation-based Evaluation, we evaluate the performance of GLMM tree in simulated datasets, and compare it with that of GLM-based MOB and GLMMs with pre-specified interaction effects. The comparison with GLM-based MOB allows for assessing the extent to which including random-effects estimation improves the performance of MOB trees. The comparison with GLMMs with pre-specified interactions will allow for assessing the extent to which using a tree-based method will improve detection of treatment-subgroup interactions. In the Application, we apply the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. Finally, in the Discussion we summarize the results and point out some directions for future research. But first, we will introduce an artificial motivating dataset, which will be used to illustrate treatment-subgroup interaction detection with GLM-based MOB and GLMM tree.

@@

We created a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the post-treatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

The simulated dataset has 3 subgroups with differential treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment~1 is more effective than Treatment~2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment~2 is more effective than Treatment~1.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a simple random intercept, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$. For simplicity we only employ a cluster-specific intercept here, but further random effects can be easily included in $z_i$. Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree. In Study I, we assessed the accuracy of the two tree algorithms in datasets with treatment-subgroup interactions. In Study II, we assessed the Type-I error of the two tree algorithms in datasets without treatment-subgroup interactions. In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

In all simulation studies, the following data-generating parameters were varied:

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
\item Two levels for the intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
\end{enumerate}

Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. Means for the other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.

@@

In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$). Therefore, an extra level was added to the sixth facet of the \textit{General simulation design}:

\begin{enumerate}
\item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.  
\end{enumerate}

Also, to assess the effect of treatment-effect differences in nodes 3 and 7, an additional facet was added:

\begin{enumerate}
\item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

50 datasets were generated for each cell of the design, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated. In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.

@@

For the first split, GLMM tree selected the true partitioning variable ($U_2$) in all datasets, and GLM tree in all but one datasets. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}).

@@

The total number of nodes in estimated GLM and GLMM trees were calculated and a tree of size $>1$ was classified as a Type-I error. To identify the most important predictors of Type-I error, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors were further assessed with a graphical display.

@@

The graphical display in Figure~\ref{fig:xyplot_treesize_nointeract} indicates that GLMM tree has a Type-I error rate somewhat below the pre-specified $\alpha$ level under all circumstances. The same goes for GLM tree, whenever the random intercept has 0 variance, or the random intercept is not correlated to one of the partitioning covariates. However, when the random intercept is correlated with one of the potential partitioning covariates, the type-I error rapidly increases for GLM tree. With large sample sizes and large variance of the random intercept, GLM tree will almost certainly create spurious splits.

@@

For datasets with purely continuous interactions, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$. Here, $\beta$ has a global value and no subscript, comprising purely continuous main and (treatment) interaction effects of the moderator variables $U$, as described in Table \ref{tab:continuous_terms}.

@@

\subsubsection{GLMMs with pre-specified interactions}
GLMMs were estimated by specifying main effects for all covariates $U_k$ and the treatment indicator, first-order interactions between all pairs of covariates $U_k$, and second-order interactions between all pairs of covariates $U_k$ and treatment. Continuous predictor variables were centered by subtracting the mean value, before calculating and including the interaction term in the GLMM.

\subsubsection{Assessment of performance} 
Predictive accuracy was assessed in terms of the correlation between the true and predicted treatment-effect differences in test datasets. As the full GLMM models were likely to overfit, predictions for GLMMs with pre-specified interactions were obtained by refitting a GLMM on the training data, using only the predictors with p-values $< .05$ in the original GLMM. Predictions for test observations were obtained from these refitted GLMMs.

To identify the most important predictors of predictive accuracy, we used an ANOVA with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors were further assessed with a graphical display.

@@

GLMM tree showed somewhat higher average correlations between the true and predicted treatment-effect differences: average accuracy was .54 (SD = .40) for GLMM trees and .51 (SD = .43) for GLMMs. The most important predictors of predictive accuracy were further assessed by means of the graphical display depicted in Figure~\ref{fig:xyplot_correlation_pw}.

As Figure~\ref{fig:xyplot_correlation_pw} indicates, GLMM trees show highest predictive accuracy in datasets with purely piecewise interactions, whereas GLMMs show highest predictive accuracy in datasets with purely continuous interactions. GLMMs perform poorly when interactions are not purely piecewise, and GLMM tree performace poorly when interactions are purely linear.

Performance of both methods improves with increasing sample size. Furthermore, performance of GLMM tree is not affected by the number of covariates, whereas the predictive accuracy of GLMMs deteriorates when the number of covariates increases, especially when the true interactions are not purely continuous. This indicates that GLMM tree may be especially useful for exploratory purposes, where there are many potential moderator variables.

GLMM trees may also provide simpler models, as the GLMMs included 12.30 significant predictor variables (main effects and interactions), on average. GLMM trees had 3.38 inner nodes on average, requiring less variables to be evaluated for making predictions.

@@

In the current paper, we presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. As such, we hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies. Also, we illustrated the application of GLMM tree on an existing dataset.

@@

GLMM tree also provided better predictive accuracy than GLM tree, with an average correlation between true and predicted treatment differences of .94 for GLMM tree, and .88 for GLM tree. GLMM tree clearly outperformed GLM tree when random effects in the datasets were sizable, differences in treatment effects across terminal nodes were relatively small (Cohen's $d=.5$), and/or sample size was small ($< 1,000$). Such treatment-effect differences and sample sizes may be quite common, even in multi-center clinical trials, and GLMM tree may provide a helpful tool for subgroup detection in those instances.

@@

Compared to treatment-interaction detection by means of GLMMs with pre-specified interaction effects, GLMM trees provided somewhat better accuracy, on average. GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at leas partially piecewise. Our findings also indicated a clear advantage for GLMM tree when there are a large number of potential moderator variables (i.e., $> 5$). Therefore, GLMM trees may be better suited than GLMMs for exploratory analyses, in which moderator variables need to be selected from a larger number of covariates. Furthermore, the number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models. The trees in our simulations were limited to a maximum number of 7 inner nodes, and may therefore be much easier to use for prediction in practical decision-making contexts.

@@

In the Application, we found GLMM tree to provide results that were more easily interpretable, and also more accurate than a GLM tree without random effects. In addition, to judge clinical relevance of the findings, we calculated node-specific effect sizes, as would often be done in RCTs or meta-analysis. Although we have limited ourselves to calculating Cohen's $d$ in the current paper, equivalent values of the success rate difference or the number needed to treat can be calculated, but this would involve additional distributional assumptions \cite{KraeyKupf06}. Node-specific effect sizes can also be used to prune trees, when a researcher prefers to have a final tree which is based on statistical as well as clinical significance. A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significance in the tree-growing process.
